Machine Learning Autoencoder Applied to Communication Channels

### 2019-06-25 Rapport/Article/Article.tex
Seance 06/25/2019

\section{Introduction}
\subsection{Motivation}
A point to point communication channel is a system in which  two terminals exchange information through a noisy channel as Fig. \ref{fig:cs} illustrates. As a result of channel imperfections, Shannon theorized in \cite{Shannon:2001:MTC:584091.584093} the ultimate reliable data rate which can be transmitted through a communication system with an arbitrarily small probability error.

@@

Recently, a significant amount of work in radio communication theory has emerged, introducing ML elements to the communication system. O'Shea et al. in \cite{2016arXiv160806409O} developed a channel autoencoder with optimized impairment and regularization layers to emulate channel impairments. They studied this architecture over an additive white Gaussian noise (AWGN) channel, founding ``some promising initial capacity" for this scheme. In their research, results in terms of BER over SNR for a DNN based autoencoder and for a convolutional neural network (CNN) based autoencoder were treated. They used a range of SNR - from $-10 \text{dB}$ to $15 \text{dB}$ - with QPSK and QAM16 modulation as benchmarks. This analysis was conducted for a binary input message.

T. Gruber et. al. in \cite{2017arXiv171008379G} proved that a deep learning-based channel decoder could actually learn a decoding algorithm rather than just being a simple classifier. They introduced code words that were not been used in the training set, and the trained NN was able to correctly decode it. They also observed that structured codes are easier to learn than unstructured ones. NN for structured codes are able to generalize to the full codebook even if they have not seen all the training examples. They trained the NN for very short blocklengths ($N \leq 64$) in order to compare with MAP decoding performance.

@@

In this work, we will implement a ML autoencoder for a BSC which performs similar to the MAP decoder in real applications for a range of cross over probability $p$ from $0$ to $0.1$. In its most simple form, a channel autoencoder includes an encoder, a noisy channel and a decoder. Using state-of-the art DNN algorithms to find the best solution of the problem, this work aims to contribute to set up a higher standard in terms of performance in bit-error correction and reduced delay for digital communication applications. In a near future, this disruptive methodology for error correction using ML could replace mathematically optimal decoders which are the current guideline.

@@

The main theoretical concepts will be briefly introduced. They are organized by theme and is aimed to a reader less familiar with the particular subject.

@@

In particular, linear block codes represented by a pair $(n, k)$ called \textit{code name} will be implemented. Where $k$ is the length of a source message and $n$ is the length of a code word. Briefly, a linear encoder will index a message $\textbf{u}^k \in \mathcal{U} \subseteq \{0,1\}^k$ to a code word $\textbf{x}^n \in \mathcal{X} \subseteq \{0,1\}^n$. This mapping is done by the \textit{generator matrix} $ \textbf{G} $ through equation \ref{eq:encoding}.

@@

The Hamming distance $d_H:\mathcal{X}\mapsto \mathbb{Z}$ is a metric in the word set which in the context of linear coding is equal to equation \ref{eq:dh}. Noteworthy for the MAP rule, a non zero Hamming distance indicates that there is a block error, and ultimately at least one binary error in the received code word.

@@

Turbo decoding using Log-MAP decoder for an additive white Gaussian noise (AWGN) channel and for a BSC was studied in \cite{JordanMA}. They analyzed the BER for small SNR and concluded that the results are sensible to SNR estimation of the real channel, since the estimation of this parameter is required for the metric calculation of the algorithm \cite{journals/ett/RobertsonHV97}. It means that if the real SNR is smaller, the decoder will not perform well.

@@

\caption{MLNN representative diagram, where $\textbf{y}^n$ is the input vector, $\textbf{r}_{l}^{j}$ is a hidden layer vector
and $\hat{\textbf{u}}^{k}$ is the output vector.} \label{fig:NN}
\end{figure}

@@

For a classification problem in communication, the \textit{categorical cross-entropy} loss function $l(\textbf{p},\textbf{q}):\mathbb{R} ^{N_L}\times \mathbb{R} ^{N_L}\mapsto \mathbb{R}$ defined in equation \eqref{eq:c-e} is most common. It is derived from the log-likelihood of a training set where $q_j$ is the estimated probability of the outcome $j$ and $p_{j}$ is the true probability. Basically, the function measures the dissimilarity between $p_{j}$ (what you expected) and $q_{j}$ (what you obtained) \cite{murphy2013machine} where $j=1,...,N_L$. Another commonly used loss function, indispensable for this research, is the \textit{mean squared error} (MSE). The MSE is the  sum of squared distances between our target variable and predicted values and it is defined in equation \ref{eq:mse}.

@@

With all these elements set, training the neural network to calculate an accurate weight matrix $\textbf{W}$ for forward propagation requires a labeled \textit{training data set}. This set is composed of pairs $ (\textbf{y}_{i}^{n}, {\mathbf{u}}_i^k) $, where $i=1,...,S$ and $S$ is the number of training sets. It matches the input and its respective desired output. The objective is to minimize the overall loss function defined in \eqref{eq:o-c-e} in terms of the parameter $\theta$ \cite{DBLP:journals/corr/OSheaH17}.

@@

For calculating the BER versus the crossover probability $p$ with confidence over the results, a Monte Carlo simulation was elaborated. This stochastic computational algorithm uses pseudo-random sampling to solve different problems. Hence, we can work with a wide variety of transmission scenarios of the communication system.

Applied to the context of this research, the messages are repeatedly randomly generated. They are coded by a generator matrix or by a DNN encoder and then transmitted through the BSC channel of crossover probability $p$. Their information arrives corrupted by noise in the decoder and is subsequently partially corrected by a decoder. Finally, the average BER is calculated and plotted. The interval of values of $p$ is in between $0$ and $0.10$.

@@

Although the MAP algorithm is the mathematically optimal decoder as cited previously, it has a high time complexity $(\mathcal{O}(2^k))$, since it computes the Hamming distance between the received code word and every code word from the code book. For more details, the pseudo code of the algorithm is available below.

@@

The architecture of the DNN decoder necessary to achieve a great performance in terms of BER is not evident. To begin, we must define which layers were used and why. A \textit{Lambda} layer is a layer in which the activation function is user defined and the parameters are non-trainable. Thus, we were able to define a layer that act as a BSC to force our DNN to learn its effects in the training data. The fact that the parameters are non-trainable means that the noise layer is applied in forward propagation only.

A Dense layer is a densely connected layer, which means that each neuron receives input from all the neurons in the previous layer. Thus it is often used to change the dimensions of a vector. In other words, it applies a rotation, scaling and translation transform to a vector as shown in equation \ref{eq:eqFP}. The values in the weight matrix are trainable parameters which are updated during BP \cite{chollet2015keras}.

The choice of the crossover probability of the training channel, $p_{t}$, was based in the work of \cite{DBLP:conf/acssc/BenammarP18} which shows $p_{t}=0.07$ to be the optimal in terms of decoding precision after the DNN is trained. In addition, a deeper DNN showed best results in \cite{DBLP:journals/corr/OSheaH17} where 3 hidden layers were used. Consequently, the same architecture were adopted to solve the problem as shown in table \ref{tab:arraydecoder}.

@@

The training set contained all the possible messages as training labels and its code words as training data. Because of the chosen encoding, the training data is a $256\times16$ matrix and the training label is a $256\times8$ matrix. To generate a message, a stochastic algorithm was developed to fabricate random 8-length vectors of integers 1s and 0s which the average value considering all elements was around $0.5$. In this way, we could guarantee that our training environment were representative of a real life application. Where different bits are transmitted with the same probability.

Even though it is not recommended to train a NN with all the available data - because it might lead to over-fitting -, for this case does not seems to be a problem. Apparently, the presence of a BSC is enough to regularize the training procedure and prevent over-fitting.

The choice of training parameters is a mix between what is standardized in literature and a hyper parametric analytic approach, where several trials lead to a decision. First, the binary cross-entropy as loss function is widely used in literature and showed to converge to the desired result consistently during tests. For the optimizer, the \textit{Adam} built-in Keras optimizer was chosen. The Adam optimizer algorithm is similar to a classic stochastic gradient descent algorithm, but computes individual adaptive learning rates for different parameters from estimates of first and second moments of the gradients \cite{chollet2015keras}. Next, a batch size matching the size of the training set which is sufficiently small to still operate in the regime of small batch was chosen. In \cite{DBLP:journals/corr/KeskarMNST16}, the authors demonstrated better performance for a NN training with stochastic gradient descent and its variants operating in a small-batch regime, usually between 32-512 data points. Finally, the necessary number of epochs to training is sensitive to the problem, since it impacts directly the training computational cost. As a result, we tried several training epoch sizes and found $2^{16}$ to have precise results and be computationally viable. The training parameters are displayed in table \ref{tab:arraydecoderTrain}.

@@

The architecture of the one-hot decoder is much simpler when compared to the array decoder. It contains only one hidden layer of $256$ units which maps the input of size 16 to the $2^8=256$ possible messages. Hence, this characterizes a multi-class classification problem, meaning that the chosen activation function is the Softmax (Eq. \ref{eq:soft}). This training technique is known in the literature as \textit{one-hot} training.

The denomination one-hot comes from the output vector which is usually a long vector of elements close to zeros with one element close to one. This \textit{hot} vector element indicates which class the NN has classified the input with the most probability. In this context, which message encoded in 256 different classes the code word refers to. Basically, a message is mapped into a class and a class only and the decoder determines this class based on a code word input.

@@

Despite the NN complexity benefits of the one-hot decoder, it is required to transform the predicted vector into a message. Consequently, an additional operation is needed. To transform from a one-hot vector to a message, a ordered collection of the messages was stored in memory. The target message is determined by a $argmax$ operation in the one-hot vector, which identifies which message is the most likely to have been transmitted through the channel.

The training labels in this case differs from the array decoder. They are one-hot vectors which maps into real messages through a bijective function. Thus, it is a matrix of dimension $256\times256$ composed of 256 different on-hot vectors. The training data is still the linear encoding of the real messages by the matrix $\textbf{G}$.

The one-hot decoder was trained both with and without a noise layer. The model without a integrated channel converged faster to the solution. Tables \ref{tab:onehotdecoder} and \ref{tab:onehotdecoderTrain} resume the architecture and training variables, respectively.

@@

Even though DNN decoders are useful, an encoding matrix is still needed. Aiming to eliminate completely the need of an explicit encoding, a technique commonly used in ML for image recognition is explored. The autoencoder is able to learn not only a decoding algorithm, but also a encoding one as introduced in the following section.

\subsection{Autoencoder}
For the DNN autoencoder, two different concepts were studied. The array autoencoder takes as input the messages and it tries to replicate them. It encodea the messages into a 16 bit code word, transmits the code word through a noisy channel and decodes them in the second portion of the DNN. A one-hot autoencoder is inputted by a one-hot vector representing one unique $k$-bit message. It encodes the vector into a $n$-bit long code word, transmits them through a noisy channel and decode them in the second part of the DNN.

This end-to-end approach for both architectures is promising for digital communication, since there is no need for a predefined linear encoding. Theoretically, the DNN could learn the best encoding and associated decoding to a specific channel. Once trained, they could perform optimally in terms of encoding and decoding for a channel.

To be clear in the discussion of this and the following sessions, the autoencoder will be divided in three: the encoder, the channel and the decoder portions. Fig. \ref{fig:DDNNAutoencoder} illustrates this division. Even though the autoencoder could in theory be able to achieve at least the same performance as the MAP, finding the parameters which unveils its capabilities requires multiple testing and a rigorous hyper-parametric analysis. In total, there were tested 71 array and 80 one-hot autoencoders different architectures or training configurations.

@@

For the array autoencoder, we first experimented with the depth of the encoder. The number of hidden layers in a DNN influences the BP algorithm performance. Extremely deep NN, with more than 4 hidden layers, may encounter problems in the gradient calculations due to vanishing weights. When the weights becomes too small, they may be source of underflow, depending on the machine precision.

The encoder portion was first analyzed in a standalone session. We trained the encoder to replicate the linear encoding proposed in the preceding sections as we studied the effects of each parameter in the training performance. A more wide architecture showed faster convergence and the use of three or more hidden layers had no positive effects. In the end, for the integrated model, we opted for using two hidden layers, with 512 and 256 densely connected units as detailed in table \ref{tab:arrayautoencoder}. The activation functions remained the same as in the DNN decoders discussed in the previous section.

@@

The decoder architecture was also based on the standalone array decoder. In spite of this, there were no outstanding difference in decoding with or without the last hidden layer of the decoder. In the other hand, the tests with a condensate NN with one hidden layer demanded a greater training time to arrive in the same results. Hence, we dropped one hidden laeyer from the cedoder and kept two

The final test was done with the entire autoencoder and the final tweaks were done. While training the autoencoder, we noticed that the binary cross-entropy loss function would need more training epochs than the MSE and \textit{logcosh} loss functions, even though they converged in similar results. The batch size was maintained as the size of the first dimension of the training set and there was no reason for changing the optimizer. Notwithstanding, the ensemble of changes and optimizations were not enough to find a suitable architecture for channels with $p<0.02$ as the MAP algorithm for the 1/2 rate encoder would do better in this scenario. Table \ref{tab:arrayautoencoderTrain} summarizes the array autoencoder training parameters.

Furthermore, adding batch normalization layers inside the DNN was fundamental. With greater training parameters, achieving convergence in a reasonable time requires more computing power of no specific technique is implemented. The frequent changes of weight values in the intermediate layers during each training step   is the principal contributor to the training cost. This problem is known as \textit{internal covariance shift}. What batch normalization does is normalize the input of a layer by its batch statistics and then scale and shift it. For further details, refer to \cite{DBLP:journals/corr/IoffeS15} which explain their usability in demanding training scenarios.

@@

@@

This section is divided in three parts. The first shows the results obtained for the two decoders. Secondly, the results for the autoencoders is brought with some observations during the training period. The discussion and analysis of these two subsections is around the BER performance. Finally, a decoding time performance analysis is done and the results are shown and discussed.

@@

The MAP decoder and the DNN decoders were tested with the same number of messages and channels. As a result, since they express the same results, we can assume that the DNN decoder was able to learn the MAP algorithm during training phase. More interestingly, is the difference between the convergence period of the decoders. The one-hot decoder reached its peak performance much faster in comparison with the array decoder in terms of training epochs. We would attribute this increased performance to the activation function of the last layer and the implementation of the BSC in the first decoder. The sigmoid function used in the array decoder may lead to sharp damp gradients, gradient saturation or slow convergence during BP. While the one-hot decoder uses the softmax function which its partial derivatives have a better behavior during BP. Refer to \cite{2018arXiv181103378N} for deeper details in each activation function characteristics. And the noise drawback in training a NN is enlarging its training time.

@@

In order to analyze the delay in the communication for each technique studied, a Monte Carlo simulation was conducted. There were selected 20 different BSC channels. For each channel, $100000$ messages were codified, transmitted and decodified for each decoder, autoencoder and MAP. This process was timed and repeated several times. In the end, an average time and standard deviation of the results were recorded. A normalization for the average MAP decoding time was done. The results are available in the table \ref{tab:timeanalysis}.

@@

Although the one-hot decoder has considerably fewer parameters than the array decoder, it must calculate a $argmax$ function to deliver the actual message that was transmitted. This extra costly computation is also the reason for the increased time for the one-hot autoencoder. It must first encode the message, then transmit it and finally decode it, performing two times per message a $argmax$ operation.

@@

We also proved that autoencoders can learn a non linear encoding function and its decoding function for a binary symmetric channel. Thus there is no need for a generator matrix for error correction. There are still some challenges in training to solve, such as ameliorate the architecture so it can outperform the reference model in every channel condition. In addition, because of the number of parameters, their calculation in CPU is slow and they showed a greater delay when compared to the MAP rule. Though, they must take the most advantage of GPU based calculations and the delay time could be improved using this technology.

A more rigorous hyper-parametric analysis could find a combination of parameters which yields better BER performance for the autoencoders and is strongly suggested for future work. In addition, in the future, a real implementation of the DNN decoders could be tested to confirm their capabilities as it is not the main focus of this work. We suggest using more advanced NN topology such as a recurrent neural network (RNN) or a generative adversarial network (GAN). Their implementation and training are more complex, however they have recently demonstrated outstanding results in other fields.

%Talk about feasibility of decoders, improved performance of the channel delay with DNN decoders, challenges in implementing this architecture in real life, challenges in training the NNs, Future work, what else could be done with NN in error correction, Advantages and disadvantages in comparison with linear decoding.

@@

I would like to thank the guidance and advice from my supervisor Meryem Benammar, Ph.D, my colleague Rémy Zawislak who worked in the same theme and collaborated indirectly to the research and Marjorie Grzeskowiak Lucas, Ph.D, who through the Institut Supérieur de l'Aéronautique et de l'Espace (ISAE-SUPAERO) enabled the confection of the project.

>>>>

\section{Introduction}
\subsection{Motivation}
A point to point communication channel is a system in which  two terminals exchange information through a noisy channel as Fig. \ref{fig:cs} illustrates. As a result of channel imperfections, Shannon theorized in \cite{Shannon:2001:MTC:584091.584093} the ultimate reliable data rate which can be transmitted through a communication system with an arbitrarily small probability of error.

@@

Recently, a significant amount of work in radio communication theory has emerged, introducing ML elements to the communication system. O'Shea et al. in \cite{2016arXiv160806409O} developed a channel autoencoder with optimized impairment and regularization layers to emulate channel impairments. They studied this architecture over an additive white Gaussian noise (AWGN) channel, founding ``some promising initial capacity" for this scheme. In their research, results in terms of BER over SNR for a DNN based autoencoder and for a convolutional neural network (CNN) based autoencoder were treated. They used a range of SNR from $-10 \text{dB}$ to $15 \text{dB}$ with QPSK and QAM16 modulation as benchmarks. This analysis was conducted for a binary input message.

T. Gruber et. al. in \cite{2017arXiv171008379G} proved that a deep-learning based channel decoder could actually learn a decoding algorithm rather than just being a simple classifier. They introduced code words that were not been used in the training set, and the trained NN was able to correctly decode it. They also observed that structured codes are easier to learn than unstructured ones. NN for structured codes are able to generalize to the full codebook even if they have not seen all the training examples. They trained the NN for very short blocklengths ($N \leq 64$) in order to compare with MAP decoding performance.

@@

In this work, we will implement a ML autoencoder for a BSC which performs similar to the MAP decoder in real applications for a range of cross over probability $p$ from $0$ to $0.1$. In its most simple form, a channel autoencoder includes an encoder, a noisy channel and a decoder using state-of-the art DNN algorithms to find the best solution of the problem. This work aims to contribute to set up a higher standard in terms of performance in bit-error correction and reduce delay for digital communication applications. In a near future, this disruptive methodology for error correction using ML could replace mathematically optimal decoders which are the current guideline.

@@

The main theoretical concepts will be briefly introduced. They are organized by theme and are aimed to a reader less familiar with the particular subject.

@@

In particular, linear block codes represented by a pair $(n, k)$ called \textit{code name} will be implemented. Where $k$ is the length of a source message and $n$ is the length of a code word. Briefly, a linear encoder will index a message $\textbf{u}^k \in \mathcal{U} \subseteq \{0,1\}^k$ to a code word $\textbf{x}^n \in \mathcal{X} \subseteq \{0,1\}^n$. This mapping is done by the \textit{generator matrix} $ \textbf{G} $ through equation \eqref{eq:encoding}.

@@

The Hamming distance $d_H:\mathcal{X}\mapsto \mathbb{Z}$ is a metric in the word set which in the context of linear coding in digital communication is equal to equation \eqref{eq:dh}. Noteworthy for the MAP rule, a non zero Hamming distance indicates that there is a block error, and ultimately at least one binary error in the received code word.

@@

Turbo decoding using Log-MAP decoder for an additive white Gaussian noise (AWGN) channel and for a BSC was studied in \cite{JordanMA}. They analyzed the BER for small SNR and concluded that the results are sensible to SNR estimation of the real channel, since the estimation of this parameter is required for the metric calculation of the algorithm \cite{journals/ett/RobertsonHV97}. It means that, the smaller the real SNR is, the worst is the decoder performance.

@@

\caption{NN representative diagram, where $\textbf{y}^n$ is the input vector, $\textbf{r}_{l}^{j}$ is a hidden layer vector
and $\hat{\textbf{u}}^{k}$ is the output vector.} \label{fig:NN}
\end{figure}

@@

For a classification problem in communication, the \textit{categorical cross-entropy} loss function $l(\textbf{p},\textbf{q}):\mathbb{R} ^{N_L}\times \mathbb{R} ^{N_L}\mapsto \mathbb{R}$ defined in equation \eqref{eq:c-e} is most common. It is derived from the log-likelihood of a training set where $q_j$ is the estimated probability of the outcome $j$ and $p_{j}$ is the true probability. Basically, the function measures the dissimilarity between $p_{j}$ (what you expected) and $q_{j}$ (what you obtained) \cite{murphy2013machine} where $j=1,...,N_L$. Another commonly used loss function, indispensable for this research, is the \textit{mean squared error} (MSE). The MSE defined in equation \eqref{eq:mse} is the sum of the squared distances between the target variable and the predicted values.

@@

With all these elements set, training the neural network to calculate an accurate weight matrix $\textbf{W}$ for forward propagation requires a labeled \textit{training data set}. This set is composed of pairs $ (\textbf{y}_{i}^{n}, {\mathbf{u}}_i^k) $, where $i=1,...,S$ and $S$ is the number of training sets. It matches the input and its respective desired output. The objective of training is to minimize the overall loss function defined in \eqref{eq:o-c-e} in terms of the parameter $\theta$ \cite{DBLP:journals/corr/OSheaH17}.

@@

For calculating the BER versus crossover probability $p$ with confidence over the results, a Monte Carlo simulation was elaborated. This stochastic computational algorithm uses pseudo-random sampling to solve different problems. Hence, we can work with a wide variety of transmission scenarios of the communication system.

Applied to the context of this research, the messages are repeatedly randomly generated. They are coded by a generator matrix or by a DNN encoder and then transmitted through the BSC channel of crossover probability $p$. Their information arrives corrupted by noise in the decoder and is subsequently completely or partially corrected by a decoder. Finally, the average BER is calculated and plotted. The interval of values of $p$ is in between $0$ and $0.10$.

@@

Although the MAP algorithm is the mathematically optimal decoder as aforementioned, it has a high time complexity $(\mathcal{O}(2^k))$, since it computes the Hamming distance between the received code word and every code word from the code book. For more details, the pseudo code of the algorithm is available below.

@@

The architecture of the DNN decoder necessary to achieve a great performance in terms of BER is not evident. To begin, we must define which layers have been used and why. A \textit{Lambda} layer is a layer in which the activation function is user defined and the parameters are non-trainable. Thus, we were able to define a layer that act as a BSC to force our DNN to learn its effects in the training data. The fact that the parameters are non-trainable means that the noise layer is applied in forward propagation only.

A Dense layer is a densely connected layer, which means that each neuron receives input from all the neurons in the previous layer. Thus it is often used to change the dimensions of a vector. In other words, it applies a rotation, scaling and translation transform to a vector as shown in equation \eqref{eq:eqFP}. The values in the weight matrix are trainable parameters which are updated during BP \cite{chollet2015keras}.

The choice of the crossover probability of the training channel, $p_{t}$, was based in the work \cite{DBLP:conf/acssc/BenammarP18} which shows $p_{t}=0.07$ to be the optimal in terms of decoding precision after the DNN is trained. In addition, a deeper DNN showed best results in \cite{DBLP:journals/corr/OSheaH17} where 3 hidden layers were used. Consequently, the same architecture were adopted to solve the problem as shown in table \ref{tab:arraydecoder}.

@@

The training set contained all the possible messages as training labels and its code words as training data. Because of the chosen encoding, the training data is a $256\times16$ matrix and the training label is a $256\times8$ matrix. To generate a message, a stochastic algorithm was developed to fabricate random 8-length vectors of integers 1s and 0s which the average value considering all elements was around $0.5$. In this way, we could guarantee that our training environment were representative of a real life application. Where different bits are transmitted with equal probability.

Even though it is not recommended to train a NN with all the available data, because it might lead to over-fitting, for this case it does not seems to be a problem. Apparently, the presence of a BSC is enough to regularize the training procedure and prevent over-fitting.

The choice of training parameters is a mix between what is standardized in literature and a hyper parametric analytic approach, where several trials lead to a decision. First, the binary cross-entropy was used as loss function since it is widely used in literature and showed to converge to the desired result consistently during tests. For the optimizer, the \textit{Adam} built-in Keras optimizer was chosen. The Adam optimizer algorithm is similar to a classic stochastic gradient descent (SGD) algorithm, but computes individual adaptive learning rates for different parameters from estimates of first and second moments of the gradients \cite{chollet2015keras}.

Next, a batch size matching the size of the training set which is sufficiently small to still operate in the regime of small batch was chosen. In \cite{DBLP:journals/corr/KeskarMNST16}, the authors demonstrated better performance for a NN training with stochastic gradient descent and its variants operating in a small-batch regime, usually between 32-512 data points. Finally, the necessary number of epochs to training is sensitive to the problem, since it impacts directly the training computational cost. As a result, we tried several training epoch sizes and found $2^{16}$ to have precise results and still be computationally viable. The training parameters are displayed in table \ref{tab:arraydecoderTrain}.

@@

The architecture of the one-hot decoder is much simpler when compared to the array decoder. It contains only one hidden layer of $256$ units which maps the input of size 16 to the $2^8=256$ possible messages. Hence, this characterizes a multi-class classification problem, meaning that the chosen activation function is the softmax defined in equation \eqref{eq:soft}. This training technique is known in the literature as \textit{one-hot} training.

The denomination \textit{one-hot} comes from the output vector which is usually a long vector of elements close to zeros with one element close to one. This \textit{hot} vector element indicates which class the NN has classified the input with the most probability. In this context, which message encoded in 256 different classes the code word refers to. Basically, a message is mapped into a class and a class only and the decoder determines this class based on a code word input.

@@

Despite the NN complexity benefits of the one-hot decoder, it must transform the predicted vector into a message. Consequently, an additional operation is needed. To transform from a one-hot vector to a message, a ordered collection of the messages was stored in memory. The target message is determined by a $argmax$ operation in the one-hot vector, which identifies which message is the most likely to have been transmitted through the channel.

The training labels in this case differs from the array decoder. They are one-hot vectors which maps into real messages through a bijective function. Thus, it is a matrix of dimension $256\times256$ composed of 256 different one-hot vectors. The training data is still the linear encoding of the real messages by the matrix $\textbf{G}$.

The one-hot decoder was trained both with and without a noise layer. The model without an integrated channel converged faster to the solution. Tables \ref{tab:onehotdecoder} and \ref{tab:onehotdecoderTrain} resume the architecture and training variables, respectively.

@@

Even though DNN decoders are useful, an encoding matrix is still needed. Aiming to eliminate completely the need of an explicit encoding, a technique commonly used in ML for image recognition is explored. The autoencoder is able to learn not only a decoding algorithm, but also a encoding one as demonstrated in the following section.

\subsection{Autoencoder}
For the DNN autoencoder, two different concepts were studied. The end-to-end approach for both architectures is promising for digital communication, since there is no need for a predefined linear encoding. Theoretically, the DNN could learn the best encoding and associated decoding to a specific channel. Once trained, they could perform optimally in terms of encoding and decoding for a channel.

The autoencoder will be divided in three: the encoder, the channel and the decoder portions. Fig. \ref{fig:DDNNAutoencoder} illustrates this division. Even though the autoencoder could in theory be able to achieve at least the same performance as the MAP, finding the parameters which unveils its best capabilities requires multiple testing and a rigorous hyper-parametric analysis. In total, there were tested 71 array autoencoders and 80 one-hot autoencoders different architectures and training configurations.

@@

The array autoencoder takes as input the messages and tries to replicate them as output. It encodes the messages into a 16 bit code word, transmits the code word through a noisy channel and decodes them in the second portion of the DNN.

We first experimented with the depth of the encoder. The number of hidden layers in a DNN influences the BP algorithm performance. Extremely deep NN, with more than 4 hidden layers, may encounter problems in the gradient calculations due to vanishing weights. When the weights becomes too small, they may be source of underflow, depending on the machine precision.

The encoder portion was first analyzed in a standalone session. We trained the encoder to replicate the linear encoding proposed in the reference model. We studied the effects of each parameter in the training performance. A more wide architecture showed faster convergence and the use of three or more hidden layers had no positive effects. In the end, for the integrated model, we opted for using two hidden layers, with 512 and 256 densely connected units as detailed in table \ref{tab:arrayautoencoder}. The activation functions remained the same as in the DNN decoders discussed in the previous section.

@@

The decoder architecture was also based on the standalone array decoder. In spite of this, there were no outstanding difference in decoding with or without the last hidden layer of the decoder. In the other hand, the tests with a shallower NN with one hidden layer demanded a greater training time to arrive in the same results. Hence, we dropped one hidden laeyer from the cedoder and kept two

The final test was done with the entire autoencoder and the final tweaks were done. While training the autoencoder, we noticed that the binary cross-entropy loss function would need more training epochs than the MSE and \textit{logcosh} loss functions, even though they converged in similar results. The batch size and the optimizer were maintained as there was no reason for changing them. Notwithstanding, the changes and optimization were not enough to find a suitable architecture for channels with $p<0.02$ as the MAP algorithm for the 1/2 rate encoder would do better in this scenario. Table \ref{tab:arrayautoencoderTrain} summarizes the array autoencoder training parameters.

Furthermore, adding batch normalization layers inside the DNN was fundamental. With greater training parameters, achieving convergence in a reasonable time requires more computing power if no specific technique is implemented. The frequent changes of weight values in the intermediate layers during each training step   is the principal contributor to the training cost. This problem is known as \textit{internal covariance shift}. What batch normalization does is normalize the input of a layer by its batch statistics and then scale and shift it. For further details, refer to \cite{DBLP:journals/corr/IoffeS15} which explain their usability in demanding training scenarios.

@@

A one-hot autoencoder is inputted by one-hot vectors representing  unique $k$-bit message. It encodes the vectors into $n$-bit long code words, transmits them through a noisy channel and decode them in the second part of the DNN.

@@

This section is divided in three parts. The first shows the results obtained for the two decoders. The second, the results for the autoencoders is brought with some observations during the training period. The discussion and analysis of these two subsections is around the BER performance. Finally, a decoding time performance analysis is done and the results are shown and discussed.

@@

The MAP decoder and the DNN decoders were tested with the same number of messages and channels. As a result, since they express the same results, we can assume that the DNN decoder was able to learn the MAP algorithm during training phase. More interestingly, is the difference between the convergence period of the decoders. The one-hot decoder reached its peak performance much faster in comparison with the array decoder in terms of training epochs. We would attribute this increased performance to the activation function of the last layer and the implementation of the BSC in the first decoder. The sigmoid function used in the array decoder may lead to sharp damp gradients, gradient saturation or slow convergence during BP. While the one-hot decoder uses the softmax function, which its partial derivatives have a better behavior during BP. Refer to \cite{2018arXiv181103378N} for deeper details in activation function characteristics. And the drawback of using noise layers in training a NN is enlarging the training time.

@@

In order to analyze the delay in the communication for each technique studied, a Monte Carlo simulation was conducted. There were selected 20 different BSC channels. For each channel, $100000$ messages were codified, transmitted and decodified for each model. This process was timed and repeated several times. In the end, an average time and standard deviation of the results were recorded. A normalization for the average MAP decoding time was done. The results are available in the table \ref{tab:timeanalysis}.

@@

Although the one-hot decoder has considerably fewer parameters than the array decoder, it must calculate an $argmax$ function to deliver the actual message that was transmitted. This extra costly computation is also the reason for the increased time for the one-hot autoencoder. It must first encode the message, then transmit it and finally decode it, performing two times per message an $argmax$ operation.

@@

We also proved that autoencoders can learn a non linear encoding function and its decoding function for a binary symmetric channel. Thus there is no need for a generator matrix for error correction. There are still some challenges in training to solve, such as ameliorate the architecture so it can outperform the reference model in every channel condition. In addition, because of the number of parameters, their calculation in CPU is slow and they showed a greater delay when compared to the MAP rule. Though, they must take the most advantage of GPU based calculations and the delay time could be improved by this technology.

A more rigorous hyper-parametric analysis could find a combination of parameters which yields better BER performance for the autoencoders. In addition, a real implementation of the DNN decoders could be tested to confirm their capabilities in a critical scenario. We also suggest using more advanced NN topology, such as a recurrent neural network (RNN) or a generative adversarial network (GAN). Their implementation and training are more complex, however they have recently demonstrated outstanding results in other fields.

@@

I would like to thank the guidance and advice from my supervisor Meryem Benammar, Ph.D, my colleague Rémy Zawislak who worked in the same theme and collaborated indirectly to the research and Marjorie Grzeskowiak Lucas, Ph.D, who through the Institut Supérieur de l'Aéronautique et de l'Espace (ISAE-SUPAERO) enabled the realization of the project.
